---
title: "Using STvEA to analyze CODEX data"
output: rmarkdown::github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(STvEA)
```

## Read in CODEX data
Data taken from FCS files at http://welikesharingdata.blob.core.windows.net/forshare/index.html
```{r}
data("codex_balbc1")
protein_expr <- cbind(codex_balbc1[, 1:29], MHCII=codex_balbc1$MHCII)
blanks <- codex_balbc1[,grepl("blank", colnames(codex_balbc1))]
size <- codex_balbc1$size.size

# xy coordinates are relative to tile currently, make them absolute
x <- codex_balbc1$X.X
x <- floor((codex_balbc1$tile_nr.tile_nr-1)/9) * max(codex_balbc1$X.X) + x
y <- codex_balbc1$Y.Y
y <- ((codex_balbc1$tile_nr.tile_nr - 1) %% 9) * max(codex_balbc1$Y.Y) + y
z <- codex_balbc1$Z.Z
spatial <- cbind(x,y,z)
row.names(spatial) <- row.names(protein_expr)
```

### Take corner section of CODEX data
```{r}
codex_subset <- x < 3000 & y < 3000
protein_expr <- protein_expr[codex_subset,]
blanks <- blanks[codex_subset,]
size <- size[codex_subset]
spatial <- spatial[codex_subset,]
```

## Create object to hold data
The STvEA.data class conveniently handles the required data frames and matrices between function calls.
```{r}
stvea_object <- SetDataCODEX(codex_protein = protein_expr, codex_blanks = blanks, codex_size = size, codex_spatial = as.data.frame(spatial))
```

## Filter and clean protein protein_expr
```{r}
# Remove cells that are too small or large, or have too low or too high expression in the blank channels. If lower and upper limits aren't specified, quantiles are taken as the limits.
# Normalize data by the total counts per cell
stvea_object <- FilterCODEX(stvea_object, size_lim = c(1000,25000), blank_lower = c(-1200, -1200, -1200, -1200), blank_upper = c(6000,2500,5000,2500))

# Fit a Gaussian mixture model to the expression levels of each protein, new data will be the cumulative probability according to the Gaussian with the higher mean
stvea_object <- CleanCODEX(stvea_object)
```

## Cluster CODEX cells based on protein expression
This will take around 5 minutes for ~10000 cells
```{r}
# Computes the 2 dimensional UMAP embedding of the cleaned CODEX protein expression for later visualization. UMAP also returns the KNN indices with k = n_neighbors.
stvea_object <- GetUmapCODEX(stvea_object, metric = 'pearson', n_neighbors=30, min_dist=0.1, negative_sample_rate = 50)

# Use the KNN indices from UMAP to perform Louvain clustering
stvea_object <- ClusterCODEX(stvea_object, k=30)
```

## Visualize clustering and protein expression on UMAP 
```{r}
PlotClusterCODEXemb(stvea_object)
```

```{r}
PlotExprCODEXemb(stvea_object, "CD4")
```

## Assess colocalization of clusters
```{r}
cluster_adj <- AdjScoreClustersCODEX(stvea_object, k=3)
```

```{r}
protein_adj <- AdjScoreProteins(stvea_object, k=3, num_cores=8)
```
